---
title: "Step 7: Cross-Validation (LOOCV)"
subtitle: "Danish Housing Prices · Mathematical Statistics"
author: "Carlos Castillo"
date: today
format:
  pdf:
    pdf-engine: pdflatex  # El motor más rápido para PDF
    papersize: a4         # Formato estándar en Dinamarca
    number-sections: true
    keep-tex: false       # No pierde tiempo guardando archivos basura
execute:
  cache: true             # Fundamental para velocidad en R
  freeze: auto            # Evita re-renders innecesarios
  echo: true              # Muestra tu código para el profesor
  warning: false
  results: hold           # Agrupa resultados para un renderizado más limpio
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.height = 4, fig.width = 7)
```

```{r load}
#| results: hold


library(tidyverse)
dk_villa <- readRDS("../../data/dk_villa.rds")
M_final  <- readRDS("../../data/M_final.rds")

cat("Model formula:", deparse(formula(M_final)), "\n")
cat("n =", nrow(dk_villa), "\n")
```

---

# Motivation

The training error $s(M)$ measures how well the model fits the data it was trained on. If the model is overfitted — i.e. too many predictors selected to match noise; $s(M)$ will be optimistically small and predictions on new data will be poor.

**Leave-one-out cross-validation (LOOCV)** estimates prediction error on
genuinely unseen data without requiring a separate holdout set.

---

# The LOOCV Formula 

Naïvely, LOOCV requires refitting the model $n$ times. But we can follow a shortcut via the **hat matrix**:

$$\tilde{e}_i = \frac{e_i}{1 - h_{ii}}$$

where $e_i = Y_i - \hat{Y}_i$ is the ordinary residual and $h_{ii}$ is the $i$-th diagonal of the hat matrix $H = X(X^\top X)^{-1}X^\top$.

The LOOCV estimate of $\sigma$ is then:

$$s_{\text{cv}} = \sqrt{\frac{1}{n} \sum_{i=1}^n \tilde{e}_i^2}$$

**Intuition for the shortcut:** $h_{ii}$ measures how much observation $i$ "pulls" the fitted line toward itself. Dividing by $(1 - h_{ii})$ inflates the residual to approximate what it would have been if $i$ had been left out during fitting. High-leverage points ($h_{ii}$ close to 1) get inflated most.

---

# Computation

```{r loocv}
#| results: hold


h     <- hatvalues(M_final)
e     <- residuals(M_final)
e_loo <- e / (1 - h)

s_M  <- summary(M_final)$sigma
s_cv <- sqrt(mean(e_loo^2))

cat("Training error:  s(M)  =", round(s_M,  5), "\n")
cat("LOOCV error:     s_cv  =", round(s_cv, 5), "\n")
cat("Ratio s_cv/s(M)       =", round(s_cv / s_M, 4), "\n\n")

cat("Rule of thumb:\n")
cat("  ratio ~ 1.0 : model generalises well\n")
cat("  ratio >> 1.0: model is overfitting\n")
```

---

# Leverage Diagnostics

High-leverage observations ($h_{ii} > 2p/n$ where $p$ is the number of
parameters) are worth examining — they disproportionately influence the fit.

```{r leverage}
#| results: hold


p   <- length(coef(M_final))
n   <- nrow(dk_villa)
thr <- 2 * p / n

cat("Leverage threshold 2p/n =", round(thr, 4), "\n")
cat("Observations above threshold:", sum(h > thr), "\n\n")

# Top 10 highest-leverage points
top_lev <- order(h, decreasing = TRUE)[1:10]
cat("Top 10 high-leverage observations:\n")
print(data.frame(
  obs        = top_lev,
  leverage   = round(h[top_lev], 4),
  loo_resid  = round(e_loo[top_lev], 4),
  log_price  = round(dk_villa$log_price[top_lev], 4)
))
```

---

# LOOCV Residual Plot

```{r loocv-plot}
#| results: hold


plot(fitted(M_final), e_loo,
     pch = 16, cex = 0.6, col = rgb(0, 0, 0.6, 0.4),
     main = "LOOCV Residuals vs Fitted Values",
     xlab = "Fitted (log scale)", ylab = "LOOCV residual")
abline(h =  0,       lty = 1)
abline(h =  2*s_cv,  lty = 2, col = "firebrick")
abline(h = -2*s_cv,  lty = 2, col = "firebrick")

# Highlight high-leverage points
points(fitted(M_final)[h > thr], e_loo[h > thr],
       pch = 1, cex = 1.2, col = "darkorange", lwd = 2)
legend("topright",
       legend = c("Observations", "High leverage", "±2 s_cv"),
       pch    = c(16, 1, NA),
       lty    = c(NA, NA, 2),
       col    = c(rgb(0,0,0.6,0.4), "darkorange", "firebrick"),
       bty    = "n")
```

---

# Comparison: Training Error vs LOOCV Error

```{r compare-plot}
#| results: hold


# Distribution of ordinary vs LOOCV residuals
par(mfrow = c(1, 2))

hist(e, breaks = 30, prob = TRUE,
     col = "lightsteelblue", border = "white",
     main = "Ordinary residuals",
     xlab = "Residual")
curve(dnorm(x, 0, s_M), add = TRUE, col = "firebrick", lwd = 2)

hist(e_loo, breaks = 30, prob = TRUE,
     col = "lightcoral", border = "white",
     main = "LOOCV residuals",
     xlab = "LOOCV residual")
curve(dnorm(x, 0, s_cv), add = TRUE, col = "firebrick", lwd = 2)

par(mfrow = c(1, 1))
```

---

# Summary Table

```{r summary}
#| results: hold


res_tbl <- data.frame(
  Metric = c("s(M) — training error",
             "s_cv — LOOCV error",
             "Ratio s_cv / s(M)",
             "Number of parameters p",
             "n"),
  Value  = c(round(s_M, 5),
             round(s_cv, 5),
             round(s_cv / s_M, 4),
             p,
             n)
)
knitr::kable(res_tbl, caption = "LOOCV summary")
```

**Interpretation on DKK scale:** Since the model is on the log scale, multiply $s_{\text{cv}}$ by the typical price to get an approximate absolute prediction error. For a 4.5M DKK villa: $4{,}500{,}000 \times (e^{s_{\text{cv}}} - 1) \approx$ prediction error in DKK.

```{r dkk-error}
#| results: hold


median_price <- exp(mean(dk_villa$log_price))
approx_error <- median_price * (exp(s_cv) - 1)
cat("Approximate LOOCV prediction error (DKK):",
    format(round(approx_error), big.mark = ","), "\n")
```
