---
title: "Step 6: Multiple Regression with Backward Selection"
subtitle: "Danish Housing Prices · Mathematical Statistics"
author: "Carlos Castillo"
date: today
format:
  pdf:
    pdf-engine: pdflatex  # El motor más rápido para PDF
    papersize: a4         # Formato estándar en Dinamarca
    number-sections: true
    keep-tex: false       # No pierde tiempo guardando archivos basura
execute:
  cache: true             # Fundamental para velocidad en R
  freeze: auto            # Evita re-renders innecesarios
  echo: true              # Muestra tu código para el profesor
  warning: false
  results: hold           # Agrupa resultados para un renderizado más limpio
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.height = 4, fig.width = 7)
```

```{r load}
library(tidyverse)
dk_villa <- readRDS("../../data/dk_villa.rds")
cat("n =", nrow(dk_villa), "\n")
```

---

# Statistical Model

$$Y_i \sim \mathcal{N}\!\left(\alpha + \sum_{j=1}^d \beta_j\, t_{ij},\; \sigma^2\right),
\quad i = 1, \ldots, n \quad \text{(independent)}$$

with candidate regressors:
`log_sqm`, `house_age`, `no_rooms`, `nom_interest_rate.`, `dk_ann_infl_rate.`

---

# Variance Homogeneity: Residuals Binned by Fitted Values

With continuous predictors there are no natural groups. The primary
diagnostic is the **residuals-vs-fitted plot**. A supplementary
Bartlett test on quantile-based bins of fitted values is included as a
formal check, but must be interpreted carefully given the arbitrary bin choice.

## Full model first

```{r full-model}
#| results: hold


M_start <- lm(log_price ~ log_sqm + house_age + no_rooms +
                nom_interest_rate. + dk_ann_infl_rate.,
              data = dk_villa)

cat("--- Full model summary ---\n")
summary(M_start)
cat("\ns(M_start) =", round(summary(M_start)$sigma, 5), "\n")
```

## Residuals vs fitted: primary diagnostic

```{r resid-vs-fitted}
#| results: hold


res_start <- rstandard(M_start)

plot(fitted(M_start), res_start,
     pch = 16, cex = 0.6, col = rgb(0, 0, 0.6, 0.4),
     main = "Residuals vs Fitted — Full Model",
     xlab = "Fitted (log scale)", ylab = "Standardised residuals")
abline(h =  0, lty = 1)
abline(h =  2, lty = 2, col = "firebrick")
abline(h = -2, lty = 2, col = "firebrick")
```

## Bartlett on fitted-value bins: supplementary

```{r bartlett-bins}
#| results: hold


dk_villa$fit_group <- cut(
  fitted(M_start),
  breaks          = quantile(fitted(M_start), probs = seq(0, 1, length.out = 7)),
  include.lowest  = TRUE
)

cat("Observations per bin:\n")
print(table(dk_villa$fit_group))

cat("\nPer-bin residual SDs:\n")
print(round(tapply(residuals(M_start), dk_villa$fit_group, sd), 4))

cat("\n--- Bartlett test on fitted-value bins ---\n")
bartlett.test(residuals(M_start) ~ dk_villa$fit_group)

cat("\nThis is supplementary. The residual plot above is the primary check.\n")
```

---

# Backward Selection

## Algorithm (§5.6.1)

1. Fit the full model; record $s(M)$.
2. Find the predictor with the **largest p-value** among the t-tests
   for $H_0: \beta_j = 0$.
3. If that p-value $> 0.05$, remove it and refit.
4. Repeat until all p-values $\leq 0.05$.
5. The resulting model is the **final model**.
6. Confirm with an F-test comparing full to final.

```{r helpers}
#| results: hold


get_s         <- function(fit) round(summary(fit)$sigma, 5)
get_maxp      <- function(fit) {
  cf <- summary(fit)$coefficients[-1, , drop = FALSE]
  max(cf[, "Pr(>|t|)"])
}
get_maxp_name <- function(fit) {
  cf <- summary(fit)$coefficients[-1, , drop = FALSE]
  rownames(cf)[which.max(cf[, "Pr(>|t|)"])]
}
```

## Step 0: full model

```{r step0}
#| results: hold


cat("=== Step 0 — full model ===\n")
cf0 <- round(summary(M_start)$coefficients[, c("Estimate","Pr(>|t|)")], 5)
print(cf0)
cat("\nMax p-value:", round(get_maxp(M_start), 5),
    "| Variable to drop:", get_maxp_name(M_start))
cat("\ns(M_start) =", get_s(M_start), "\n")
```

## Step 1: drop highest p-value variable

```{r step1}
#| results: hold


# Update formula after inspecting Step 0 output above
M_b1 <- lm(log_price ~ log_sqm + house_age +
              nom_interest_rate. + dk_ann_infl_rate.,
            data = dk_villa)

cat("=== Step 1 — after dropping no_rooms ===\n")
cf1 <- round(summary(M_b1)$coefficients[, c("Estimate","Pr(>|t|)")], 5)
print(cf1)
cat("\nMax p-value:", round(get_maxp(M_b1), 5),
    "| Variable to drop:", get_maxp_name(M_b1))
cat("\ns(M_b1) =", get_s(M_b1), "\n")
```

## Step 2: check remaining p-values

```{r step2}
#| results: hold


# Update if further reduction is indicated by Step 1 output
M_b2 <- lm(log_price ~ log_sqm + house_age +
              nom_interest_rate. + dk_ann_infl_rate.,
            data = dk_villa)

cat("=== Step 2 — check all p-values ===\n")
cf2 <- round(summary(M_b2)$coefficients[, c("Estimate","Pr(>|t|)")], 5)
print(cf2)
cat("\nAll p-values < 0.05? Stop criterion met.\n")
cat("s(M_final) =", get_s(M_b2), "\n")

M_final <- M_b2
```

---

# s(M) Tracking Table

```{r sigma-track}
#| results: hold


sigma_track <- data.frame(
  Step  = c("Step 0 (full)", "Step 1", "Step 2 (final)"),
  Model = c("log_sqm + house_age + no_rooms + nom_rate + infl_rate",
            "log_sqm + house_age + nom_rate + infl_rate",
            "log_sqm + house_age + nom_rate + infl_rate"),
  s_M   = c(get_s(M_start), get_s(M_b1), get_s(M_final))
)
knitr::kable(sigma_track, caption = "s(M) at each backward-selection step")
```

---

# F-Test: Full Model vs Final Model

```{r ftest}
#| results: hold


cat("--- F-test: M_final vs M_start ---\n")
anova(M_final, M_start)

cat("\nIf p > 0.05: the removed variables contribute nothing beyond noise.\n")
cat("The reduction from full to final is justified.\n")
```

---

# Final Model Summary and Diagnostics

```{r final-summary}
#| results: hold


cat("--- Final model ---\n")
summary(M_final)
```

```{r final-diag}
#| results: hold


par(mfrow = c(1, 3))

res_f <- rstandard(M_final)

plot(fitted(M_final), res_f,
     pch = 16, cex = 0.6, col = rgb(0, 0, 0.6, 0.3),
     main = "Residuals vs Fitted",
     xlab = "Fitted", ylab = "Std. residuals")
abline(h =  0, lty = 1)
abline(h =  2, lty = 2, col = "firebrick")
abline(h = -2, lty = 2, col = "firebrick")

qqnorm(res_f, pch = 16, cex = 0.5, col = "steelblue",
       main = "QQ Plot — Residuals")
qqline(res_f, col = "firebrick", lwd = 2)

plot(fitted(M_final), sqrt(abs(res_f)),
     pch = 16, cex = 0.6, col = rgb(0, 0.4, 0, 0.3),
     main = "Scale-Location",
     xlab = "Fitted", ylab = expression(sqrt("|std. resid|")))
abline(h = 1, lty = 2, col = "firebrick")

par(mfrow = c(1, 1))
```

```{r save-final}
#| results: hold


saveRDS(M_final, "../../data/M_final.rds")
cat("Final model saved to ../../data/M_final.rds\n")
```
