---
title: "Step 4: Two-Way ANOVA (House Type * Area)"
subtitle: "Danish Housing Prices · Mathematical Statistics"
author: "Carlos Castillo"
date: today
format:
  pdf:
    pdf-engine: pdflatex  # El motor más rápido para PDF
    papersize: a4         # Formato estándar en Dinamarca
    number-sections: true
    keep-tex: false       # No pierde tiempo guardando archivos basura
execute:
  cache: true             # Fundamental para velocidad en R
  freeze: auto            # Evita re-renders innecesarios
  echo: true              # Muestra tu código para el profesor
  warning: false
  results: hold           # Agrupa resultados para un renderizado más limpio
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.height = 4.5, fig.width = 8)
```

```{r load}
#| results: hold


library(tidyverse)
dk_clean <- readRDS("../../data/dk_clean.rds")

# Restrict to 4 areas for a balanced comparison
dk_anova <- dk_clean %>%
  filter(area %in% c("Capital, Copenhagen", "North Zealand",
                     "East & mid jutland",  "South jutland")) %>%
  droplevels()

cat("Observations per cell (house_type × area):\n")
print(table(dk_anova$house_type, dk_anova$area))
```

---

# Statistical Model: Five-Model Hierarchy

With row factor = house type ($r = 4$ levels) and column factor = area
($s = 4$ levels):

**M1 (interaction):**
$X_{ijk} \sim \mathcal{N}(\mu_{ij}, \sigma^2)$

**M2 (additive):**
$X_{ijk} \sim \mathcal{N}(\alpha_i + \beta_j, \sigma^2)$

**M3r (row only):**
$X_{ijk} \sim \mathcal{N}(\alpha_i, \sigma^2)$

**M3c (column only):**
$X_{ijk} \sim \mathcal{N}(\beta_j, \sigma^2)$

**$M_{\infty}$ (common mean):**
$X_{ijk} \sim \mathcal{N}(\mu, \sigma^2)$

---

# Variance Homogeneity: Bartlett's Test on Cells

For two-way ANOVA the relevant groups are the **crossed cells** (type × area). The course requires `interaction()` to define them correctly — the colon
operator alone does not work for `bartlett.test`.

```{r bartlett}
#| results: hold


cat("Per-cell standard deviations:\n")
cell_sds <- tapply(dk_anova$log_price,
                   interaction(dk_anova$house_type, dk_anova$area), sd)
print(round(sort(cell_sds), 3))

cat("\nMax / Min SD ratio:", round(max(cell_sds) / min(cell_sds), 2), "\n\n")

cat("--- Bartlett's test on cells ---\n")
bartlett.test(log_price ~ interaction(house_type, area), data = dk_anova)

cat("\nSame large-n caveat as Step 3: check the ratio above.\n")
```

---

# Interaction Plot

Parallel lines indicate no interaction (additive model adequate).
Non-parallel lines suggest the area effect differs by house type.

```{r interaction-plot, fig.height=5}
#| results: hold


interaction.plot(
  x.factor     = dk_anova$area,
  trace.factor = dk_anova$house_type,
  response     = dk_anova$log_price,
  col          = c("steelblue","firebrick","forestgreen","darkorange"),
  lty          = 1, lwd = 2,
  main         = "Interaction Plot: Area × House Type",
  xlab         = "Area",
  ylab         = "Mean log(Price)",
  trace.label  = "House type",
  las          = 2
)
```

---

# Model Fitting and Tests

## Test 1: Can we remove the interaction? (M1 → M2)

$$H_0: \mu_{ij} = \alpha_i + \beta_j \text{ for all } i, j
\quad \text{vs} \quad H_1: \text{interaction present}$$

```{r m1-m2}
#| results: hold


M1_2way <- lm(log_price ~ house_type * area, data = dk_anova)
M2_2way <- lm(log_price ~ house_type + area, data = dk_anova)

cat("--- F-test: M2 vs M1 (remove interaction) ---\n")
anova(M2_2way, M1_2way)
```

## Test 2: House type effect? (M3c → M2)

$$H_0: \alpha_1 = \alpha_2 = \alpha_3 = \alpha_4
\quad \text{(no house type effect given additive model)}$$

```{r m3c-m2}
#| results: hold


M3c_2 <- lm(log_price ~ area, data = dk_anova)

cat("--- F-test: M3c vs M2 (house_type effect) ---\n")
anova(M3c_2, M2_2way)
```

## Test 3: Area effect? (M3r → M2)

$$H_0: \beta_1 = \beta_2 = \beta_3 = \beta_4
\quad \text{(no area effect given additive model)}$$

```{r m3r-m2}
#| results: hold


M3r_2 <- lm(log_price ~ house_type, data = dk_anova)

cat("--- F-test: M3r vs M2 (area effect) ---\n")
anova(M3r_2, M2_2way)
```

---

# Additive Model: Summary

```{r m2-summary}
#| results: hold


summary(M2_2way)

cat("\nResidual sigma s =", round(summary(M2_2way)$sigma, 4), "\n")
```

---

# Cell Mean Estimates (Additive Model)

```{r cell-means}
#| results: hold


# Reconstruct fitted means for each cell
grid <- expand.grid(
  house_type = levels(dk_anova$house_type),
  area       = levels(dk_anova$area)
)
grid$fitted_log <- predict(M2_2way, newdata = grid)
grid$fitted_dkk <- round(exp(grid$fitted_log))

cat("Fitted median prices (DKK) under additive model:\n")
mat <- matrix(format(grid$fitted_dkk, big.mark = ","),
              nrow = 4, ncol = 4,
              dimnames = list(levels(dk_anova$house_type),
                              levels(dk_anova$area)))
print(mat)
```

---

# Interpretation

The additive model is appropriate if the interaction test is not significant,meaning the **price premium of one house type over another is the same across all areas**. If interaction is significant, each type × area combination needs its own mean, and the simpler additive story breaks down.
