---
title: "Step 2: Two-Sample Comparison (2023 vs 2024)"
subtitle: "Danish Housing Prices · Mathematical Statistics"
author: "Carlos Castillo"
date: today
format:
  pdf:
    pdf-engine: pdflatex  # El motor más rápido para PDF
    papersize: a4         # Formato estándar en Dinamarca
    number-sections: true
    keep-tex: false       # No pierde tiempo guardando archivos basura
execute:
  cache: true             # Fundamental para velocidad en R
  freeze: auto            # Evita re-renders innecesarios
  echo: true              # Muestra tu código para el profesor
  warning: false
  results: hold           # Agrupa resultados para un renderizado más limpio
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.height = 4, fig.width = 7)
```

```{r load}
#| results: hold


library(tidyverse)
dk_villa <- readRDS("../../data/dk_villa.rds") %>%
  mutate(year = as.integer(substr(date, 1, 4)))

y2023 <- dk_villa$log_price[dk_villa$year == 2023]
y2024 <- dk_villa$log_price[dk_villa$year == 2024]

cat("n_2023 =", length(y2023), "| n_2024 =", length(y2024), "\n")
```

---

# Motivation

Danish nominal interest rates rose sharply from 2022 onward.
We ask: does this translate into a statistically detectable difference in
mean log-price between 2023 and 2024 Copenhagen Villas?

---

# Statistical Model

$$Y_{ij} \sim \mathcal{N}(\mu_j, \sigma^2),
\quad j \in \{2023, 2024\}, \quad i = 1, \ldots, n_j
\quad \text{(independent)}$$

---

# Variance Homogeneity — F-test

The course (§2, slides 4) prescribes the **F-test** for exactly $m = 2$ variance
estimates:

$$H_{0\sigma^2}: \sigma_{2023}^2 = \sigma_{2024}^2$$

$$F = \frac{S_{2023}^2}{S_{2024}^2} \sim F(n_1 - 1,\; n_2 - 1)
\quad \text{under } H_{0\sigma^2}$$

```{r var-test}
#| results: hold


cat("Per-group SDs:\n")
cat("  2023:", round(sd(y2023), 4), "\n")
cat("  2024:", round(sd(y2024), 4), "\n\n")

vt <- var.test(y2023, y2024)
print(vt)

use_equal_var <- vt$p.value > 0.05
cat("\nConclusion: using equal variances =", use_equal_var, "\n")
```

---

# Exploratory Plots

```{r plots}
#| results: hold


par(mfrow = c(1, 2))

boxplot(log_price ~ year, data = dk_villa,
        col  = c("lightsteelblue", "lightcoral"),
        main = "log(Price) by Year",
        xlab = "Year", ylab = "log(Price)")

plot(density(y2023), col = "steelblue", lwd = 2,
     xlim = range(dk_villa$log_price),
     main = "Density: 2023 vs 2024", xlab = "log(Price)")
lines(density(y2024), col = "firebrick", lwd = 2)
legend("topright", legend = c("2023","2024"),
       col = c("steelblue","firebrick"), lwd = 2, bty = "n")

par(mfrow = c(1, 1))
```

---

# Two-Sample t-Test

## Hypothesis

$$H_0: \mu_{2023} = \mu_{2024}
\quad \text{vs} \quad
H_1: \mu_{2023} \neq \mu_{2024}$$

## Test statistic (pooled variance)

$$T = \frac{\bar{Y}_1 - \bar{Y}_2}{S_p\sqrt{1/n_1 + 1/n_2}}
\sim t(n_1 + n_2 - 2)
\quad \text{where } S_p^2 = \frac{(n_1-1)S_1^2 + (n_2-1)S_2^2}{n_1+n_2-2}$$

```{r ttest-manual}
#| results: hold


n1  <- length(y2023);  n2  <- length(y2024)
sp2 <- ((n1-1)*var(y2023) + (n2-1)*var(y2024)) / (n1 + n2 - 2)
df  <- n1 + n2 - 2

t_stat <- (mean(y2023) - mean(y2024)) / sqrt(sp2 * (1/n1 + 1/n2))
p_val  <- 2 * pt(-abs(t_stat), df = df)
ci     <- (mean(y2023) - mean(y2024)) +
           c(-1, 1) * qt(0.975, df) * sqrt(sp2 * (1/n1 + 1/n2))

cat("mean_2023   =", round(mean(y2023), 4), "\n")
cat("mean_2024   =", round(mean(y2024), 4), "\n")
cat("sp          =", round(sqrt(sp2),   4), "\n")
cat("t-statistic =", round(t_stat, 4), "on", df, "df\n")
cat("p-value     =", round(p_val,  6), "\n")
cat("95% CI for (mu_2023 - mu_2024): [", round(ci, 4), "]\n")
cat("Back-transformed ratio CI: [",
    round(exp(ci), 4), "]\n")
```

```{r ttest-builtin}
t.test(y2023, y2024, var.equal = use_equal_var)
```

---

# Normality Check per Group

```{r normality}
#| results: hold


par(mfrow = c(1, 2))

qqnorm(y2023, main = "QQ — 2023",
       pch = 16, cex = 0.6, col = "steelblue")
qqline(y2023, col = "firebrick", lwd = 2)

qqnorm(y2024, main = "QQ — 2024",
       pch = 16, cex = 0.6, col = "steelblue")
qqline(y2024, col = "firebrick", lwd = 2)

par(mfrow = c(1, 1))
```

---

# Interpretation

The back-transformed CI for $(\mu_{2023} - \mu_{2024})$ is on the log scale. Exponentiating gives a ratio: $\exp(\bar{Y}_{2023} - \bar{Y}_{2024})$ is the multiplicative factor by which the 2023 median price exceeds the 2024 median. A value above 1 means prices were higher in 2023; below 1 means lower.
