---
title: "Step 1: One-Sample Inference on Mean Log-Price"
subtitle: "Danish Housing Prices · Mathematical Statistics"
author: "Carlos Castillo"
date: today
format:
  pdf:
    pdf-engine: pdflatex  # El motor más rápido para PDF
    papersize: a4         # Formato estándar en Dinamarca
    number-sections: true
    keep-tex: false       # No pierde tiempo guardando archivos basura
execute:
  cache: true             # Fundamental para velocidad en R
  freeze: auto            # Evita re-renders innecesarios
  echo: true              # Muestra tu código para el profesor
  warning: false
  results: hold           # Agrupa resultados para un renderizado más limpio
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.height = 4, fig.width = 7)
```

```{r load}
library(tidyverse)
dk_villa <- readRDS("../../data/dk_villa.rds")
cat("n =", nrow(dk_villa), "\n")
```

---

# Statistical Model

We model the log-prices of Copenhagen Villas (2023–2024) as:

$$Y_i \sim \mathcal{N}(\mu, \sigma^2), \quad i = 1, \ldots, n \quad \text{(independent)}$$

With a single sample there is only one variance — no homogeneity test is
applicable. The only distributional assumption to verify is normality.

---

# Normality Check

```{r normality}
#| results: hold


par(mfrow = c(1, 2))

hist(dk_villa$log_price,
     breaks = 30, prob = TRUE,
     main   = "log(Price) — histogram",
     xlab   = "log(Price)",
     col    = "lightsteelblue", border = "white")
curve(dnorm(x, mean(dk_villa$log_price), sd(dk_villa$log_price)),
      add = TRUE, col = "firebrick", lwd = 2)

qqnorm(dk_villa$log_price,
       main = "QQ plot — log(Price)",
       pch = 16, cex = 0.6, col = "steelblue")
qqline(dk_villa$log_price, col = "firebrick", lwd = 2)

par(mfrow = c(1, 1))
```

---

# Hypothesis Test

We test whether the mean log-price is consistent with a benchmark of
4,500,000 DKK:

$$H_0: \mu = \log(4{,}500{,}000) \approx 15.32
\quad \text{vs} \quad
H_1: \mu \neq \log(4{,}500{,}000)$$

Under $H_0$ the test statistic is:

$$T = \frac{\bar{Y} - \mu_0}{S / \sqrt{n}} \sim t(n-1)$$

```{r test-manual}
#| results: hold


mu0  <- log(4500000)
n    <- nrow(dk_villa)
ybar <- mean(dk_villa$log_price)
s    <- sd(dk_villa$log_price)
df   <- n - 1

t_stat <- (ybar - mu0) / (s / sqrt(n))
p_val  <- 2 * pt(-abs(t_stat), df = df)
ci     <- ybar + c(-1, 1) * qt(0.975, df) * s / sqrt(n)

cat("n           =", n,                    "\n")
cat("y-bar       =", round(ybar,   4),     "\n")
cat("s           =", round(s,      4),     "\n")
cat("mu0         =", round(mu0,    4),     "\n")
cat("t-statistic =", round(t_stat, 4), "on", df, "df\n")
cat("p-value     =", round(p_val,  6),     "\n")
cat("95% CI for mu:             [", round(ci, 4), "]\n")
cat("Back-transformed CI (DKK): [",
    format(round(exp(ci)), big.mark = ","), "]\n")
```

```{r test-builtin}
# Built-in verification
t.test(dk_villa$log_price, mu = mu0)
```

---

# Confidence Interval for $\sigma$

A $95\%$ CI for $\sigma$ uses the fact that
$(n-1)S^2/\sigma^2 \sim \chi^2(n-1)$:

$$\left[\sqrt{\frac{(n-1)S^2}{\chi^2_{0.975}(n-1)}},\;
         \sqrt{\frac{(n-1)S^2}{\chi^2_{0.025}(n-1)}}\right]$$

```{r ci-sigma}
#| results: hold


ci_sigma <- sqrt(c(
  (n - 1) * s^2 / qchisq(0.975, df),
  (n - 1) * s^2 / qchisq(0.025, df)
))
cat("95% CI for sigma: [", round(ci_sigma, 4), "]\n")
```

---

# Interpretation

**On the log scale:** the CI for $\mu$ gives the range of plausible mean
log-prices.

**On the DKK scale:** back-transforming via $\exp(\cdot)$ yields a CI for the **median** price (not the mean), because for log-normal $Y$:
$\exp\!\bigl(\mathbb{E}[\log Y]\bigr) = \text{median}(Y)$.

This is more informative for skewed price data, where the median is a more representative summary than the mean.
