---
title: "Danish Villa Prices: Simple Linear Regression"
author: "Carlos Castillo (ID: 202505985)"
date: today
format:
  pdf:
    pdf-engine: pdflatex  # El motor más rápido para PDF
    papersize: a4         # Formato estándar en Dinamarca
    number-sections: true
    keep-tex: false       # No pierde tiempo guardando archivos basura
execute:
  cache: true             # Fundamental para velocidad en R
  freeze: auto            # Evita re-renders innecesarios
  echo: true              # Muestra tu código para el profesor
  warning: false
  results: hold           # Agrupa resultados para un renderizado más limpio
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
dk_villa <- readRDS("../../data/dk_villa.rds")
```

# Introduction

Having cleaned and filtered the dataset to 402 villa transactions in the Copenhagen area from 2023–2024, we now build the first statistical model. We start with the simplest possible specification (a single continuous predictor) and verify all model assumptions before interpreting any results.

The model we want to fit is:

$$M_2: X_i \sim N(\alpha + \beta\, t_i,\, \sigma^2), \quad i = 1, \dots, 402$$

where $X_i = \log(\text{price}_i)$ and $t_i = \log(\text{sqm}_i)$.

However, before fitting $M_2$ we must verify three assumptions:

1. **Normality** — $\log(\text{price})$ is approximately normally distributed
2. **Linearity** — the mean of $\log(\text{price})$ depends linearly on $\log(\text{sqm})$
3. **Equal variance** — the variance $\sigma^2$ is constant across all values of $\log(\text{sqm})$

---

# Assumption 1: Normality

We examine the distribution of $\log(\text{price})$ on the villa subset.

```{r normality-hist}
hist(dk_villa$log_price,
     breaks = 30,
     prob   = TRUE,
     main   = "log(Price) — Copenhagen Villas 2023–2024",
     xlab   = "log(Price)",
     col    = "lightsteelblue",
     border = "white")
curve(dnorm(x, mean(dk_villa$log_price), sd(dk_villa$log_price)),
      add = TRUE, col = "firebrick", lwd = 2)
```

```{r normality-qq}
qqnorm(dk_villa$log_price,
       main = "QQ Plot — log(Price)",
       pch  = 16, cex = 0.6, col = "steelblue")
qqline(dk_villa$log_price, col = "firebrick", lwd = 2)
```

**Comment:** The histogram of $\log(\text{price})$ is approximately
bell-shaped and the QQ plot follows the theoretical normal line across the central range, with minor deviations in the tails. The normality assumption is reasonably satisfied for this subset.

---

# Assumption 2: Linearity

We test whether the mean of $\log(\text{price})$ depends **linearly** on $\log(\text{sqm})$, against the more general alternative that it depends on $\log(\text{sqm})$ in some arbitrary way.

We compare two nested models:

$$M_1: X_{ij} \sim N(\mu_i,\, \sigma^2) \quad \text{(one mean per sqm group)}$$
$$M_2: X_{ij} \sim N(\alpha + \beta\, t_i,\, \sigma^2) \quad \text{(linear regression)}$$

Since $M_2$ is a special case of $M_1$, we use an F-test:

$$H_0: \text{the mean of } \log(\text{price}) \text{ is linear in } \log(\text{sqm})$$

```{r linearity-plot}
plot(dk_villa$log_sqm, dk_villa$log_price,
     pch  = 16, cex = 0.6,
     col  = rgb(0, 0, 0.6, 0.4),
     main = "log(Price) vs log(sqm) — Copenhagen Villas",
     xlab = "log(sqm)",
     ylab = "log(Price)")

# Group means
# Try fewer breaks
# Try fewer breaks
dk_villa$sqm_group <- cut(dk_villa$log_sqm, breaks = 6)
table(dk_villa$sqm_group)

M1 <- lm(log_price ~ sqm_group, data = dk_villa)
M2 <- lm(log_price ~ log_sqm,   data = dk_villa)
anova(M1, M2)
means <- tapply(dk_villa$log_price, dk_villa$sqm_group, mean)
mids  <- tapply(dk_villa$log_sqm,   dk_villa$sqm_group, mean)
lines(mids, means, col = "firebrick", lwd = 2)
points(mids, means, col = "firebrick", pch = 16, cex = 1.2)
```

```{r linearity-test}
# M1: group means model
M1 <- lm(log_price ~ sqm_group, data = dk_villa)

# M2: linear regression
M2 <- lm(log_price ~ log_sqm, data = dk_villa)

# F-test: is the linear model adequate?
anova(M2, M1)
```
```{r}
# Use quantile-based breaks so groups are equally populated
dk_villa$sqm_group <- cut(dk_villa$log_sqm,
                          breaks = quantile(dk_villa$log_sqm,
                                           probs = seq(0, 1, length.out = 7)),
                          include.lowest = TRUE)
table(dk_villa$sqm_group)

# Refit
M1 <- lm(log_price ~ sqm_group, data = dk_villa)
M2 <- lm(log_price ~ log_sqm,   data = dk_villa)

# Correct order: larger model first
anova(M1, M2)
bartlett.test(dk_villa$log_price ~ dk_villa$sqm_group)
```


**Comment:** The group means (red points) follow a roughly linear
trend across the sqm groups. The F-test compares the linear model
$M_2$ against the unrestricted group-means model $M_1$ — if we
cannot reject $M_2$, the linearity assumption is justified.

---

# Assumption 3: Equal Variance

We test whether the variance of $\log(\text{price})$ is the same across
all sqm groups:

$$H_{0,\sigma^2}: \sigma_1^2 = \sigma_2^2 = \cdots = \sigma_8^2$$

```{r bartlett}
bartlett.test(dk_villa$log_price ~ dk_villa$sqm_group)
```

We also inspect the residual plot from $M_2$:

```{r residual-plot}
res_M2 <- rstandard(M2)
s_M2   <- summary(M2)$sigma

plot(fitted(M2), res_M2,
     pch  = 16, cex = 0.6,
     col  = rgb(0, 0, 0, 0.3),
     main = "Residuals vs Fitted — Simple Regression",
     xlab = "Fitted values",
     ylab = "Standardized residuals")
abline(h =  0, lty = 1)
abline(h =  2, lty = 2, col = "firebrick")
abline(h = -2, lty = 2, col = "firebrick")
```

**Comment:** Write your interpretation of the Bartlett result and
residual plot here after running the code.

---

# Fitting the Linear Model

Having checked the assumptions, we now fit $M_2$ and examine the
estimates.

```{r fit}
M2      <- lm(log_price ~ log_sqm, data = dk_villa)
sum_M2  <- summary(M2)
sum_M2
```

## Confidence intervals for $\alpha$, $\beta$, and $\sigma$

```{r ci}
# CI for alpha and beta
confint(M2)

# CI for sigma
s2    <- sum_M2$sigma^2
f_02  <- sum_M2$df[2]

CI_sigma <- sqrt(c(s2 * f_02 / qchisq(0.975, f_02),
                   s2 * f_02 / qchisq(0.025, f_02)))
cat("95% CI for sigma: [", round(CI_sigma, 4), "]\n")
```

## Test: is the price-size elasticity equal to 1?

A slope of $\beta = 1$ would mean that doubling the size exactly doubles
the price — a proportional relationship. We test:

$$H_0: \beta = 1 \quad \text{vs} \quad H_1: \beta \neq 1$$

```{r slope-test}
beta_hat <- sum_M2$coefficients["log_sqm", "Estimate"]
beta_se  <- sum_M2$coefficients["log_sqm", "Std. Error"]
df_02    <- sum_M2$df[2]

t_stat <- (beta_hat - 1) / beta_se
p_val  <- 2 * (1 - pt(abs(t_stat), df = df_02))

cat("beta_hat:  ", round(beta_hat, 5), "\n")
cat("t-stat:    ", round(t_stat,   5), "\n")
cat("p-value:   ", round(p_val,    6), "\n")
cat("95% CI for beta:", round(confint(M2)["log_sqm",], 5), "\n")
```

**Comment:** Write your interpretation here — is the elasticity equal
to 1? What does this mean practically for a buyer?

---

# Prediction

We use the fitted model to predict the price of a specific villa — a
130 m² property, which is close to the median size in the dataset.

```{r prediction}
new_house <- data.frame(log_sqm = log(130))

ci_mean <- predict(M2, new_house, interval = "confidence")
ci_pred <- predict(M2, new_house, interval = "prediction")

# Back-transform to DKK
cat("=== 130 sqm villa — Copenhagen 2023/2024 ===\n\n")

cat("Confidence interval for the mean price:\n")
cat("  [", round(exp(ci_mean[, "lwr"])), "-",
         round(exp(ci_mean[, "upr"])), "] DKK\n")
cat("  Point estimate:", round(exp(ci_mean[, "fit"])), "DKK\n\n")

cat("Prediction interval for a single house:\n")
cat("  [", round(exp(ci_pred[, "lwr"])), "-",
         round(exp(ci_pred[, "upr"])), "] DKK\n")
```

**Comment:** The prediction interval is wider than the confidence
interval because it accounts for two sources of uncertainty: the
uncertainty in estimating the mean (shared with the CI) and the
individual scatter of a single observation around that mean. Even
as $n \to \infty$, the prediction interval does not collapse to a
point — it converges to $\hat\alpha + \hat\beta t_0 \pm \sigma \cdot
t_{0.975}(f_{02})$, since individual houses always vary around the
mean by $\sigma^2$.

---

# Model Summary

```{r summary-table}
cat("R-squared:      ", round(sum_M2$r.squared,  4), "\n")
cat("Residual sigma: ", round(sum_M2$sigma,       4), "\n")
cat("n:              ", nrow(dk_villa),                "\n")
```

**Comment:** Write your overall conclusion here — what does R²
tell you about how well size alone explains price? What is missing?
This motivates adding more predictors in the next document.

---

# Save

```{r save}
# Save M2 for reference in subsequent documents
saveRDS(M2, "../../data/M2_simple.rds")
```
