---
title: "Step 8: Prediction Intervals for New Houses"
subtitle: "Danish Housing Prices · Mathematical Statistics"
author: "Carlos Castillo"
date: today
format:
  pdf:
    pdf-engine: pdflatex  # El motor más rápido para PDF
    papersize: a4         # Formato estándar en Dinamarca
    number-sections: true
    keep-tex: false       # No pierde tiempo guardando archivos basura
execute:
  cache: true             # Fundamental para velocidad en R
  freeze: auto            # Evita re-renders innecesarios
  echo: true              # Muestra tu código para el profesor
  warning: false
  results: hold           # Agrupa resultados para un renderizado más limpio
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.height = 4.5, fig.width = 8)
```

```{r load}
#| results: hold


library(tidyverse)
dk_villa <- readRDS("../../data/dk_villa.rds")
M_final  <- readRDS("../../data/M_final.rds")

cat("Model formula:", deparse(formula(M_final)), "\n")
cat("n =", nrow(dk_villa), "\n")
cat("s =", round(summary(M_final)$sigma, 5), "\n")
```

---

# Two Types of Interval 

Both intervals share the same point estimate $\hat{\mu}(t_0) = \hat\alpha + \hat\beta^\top t_0$
but differ in what uncertainty they capture:

**Confidence interval for the mean** — uncertainty about where the
true mean $\mu(t_0) = E[Y \mid t_0]$ lies:

$$\hat\mu(t_0) \pm t_{0.975}(n-p) \cdot s \cdot \sqrt{t_0^\top (X^\top X)^{-1} t_0}$$

**Prediction interval for a single new observation** — uncertainty about
where one new $Y_0$ will fall, adding residual variance $\sigma^2$:

$$\hat\mu(t_0) \pm t_{0.975}(n-p) \cdot s \cdot \sqrt{1 + t_0^\top (X^\top X)^{-1} t_0}$$

The prediction interval is always wider than the confidence interval by the extra $1$ under the square root, reflecting irreducible observation noise.

---

# New Houses to Predict

Three hypothetical Copenhagen Villas with characteristics representative of the 2023–2024 market:

```{r new-houses}
#| results: hold


# Adjust predictor names to match the final model's terms exactly
new_houses <- data.frame(
  log_sqm            = log(c(100, 130, 180)),
  house_age          = c(60,   50,   20),
  nom_interest_rate. = c(3.5,  3.5,  3.5),
  dk_ann_infl_rate.  = c(3.3,  3.3,  3.3)
)

new_houses$sqm <- exp(new_houses$log_sqm)
print(new_houses)
```

---

# Computing the Intervals

```{r predict-intervals}
#| results: hold


ci_log <- predict(M_final, newdata = new_houses, interval = "confidence")
pi_log <- predict(M_final, newdata = new_houses, interval = "prediction")

sqm_labels <- c("100 sqm", "130 sqm", "180 sqm")

cat("=== Confidence intervals for E[Price | t0] ===\n\n")
for (i in 1:3) {
  cat(sprintf("  %s:  point est = %s DKK\n",
      sqm_labels[i],
      format(round(exp(ci_log[i,"fit"])), big.mark = ",")))
  cat(sprintf("         95%% CI = [%s – %s] DKK\n\n",
      format(round(exp(ci_log[i,"lwr"])), big.mark = ","),
      format(round(exp(ci_log[i,"upr"])), big.mark = ",")))
}

cat("=== Prediction intervals for a single new house ===\n\n")
for (i in 1:3) {
  cat(sprintf("  %s:  95%% PI = [%s – %s] DKK\n",
      sqm_labels[i],
      format(round(exp(pi_log[i,"lwr"])), big.mark = ","),
      format(round(exp(pi_log[i,"upr"])), big.mark = ",")))
}
```

---

# Width Comparison

```{r widths}
#| results: hold


ci_widths <- exp(ci_log[,"upr"]) - exp(ci_log[,"lwr"])
pi_widths <- exp(pi_log[,"upr"]) - exp(pi_log[,"lwr"])

width_tbl <- data.frame(
  sqm          = c(100, 130, 180),
  point_est    = format(round(exp(ci_log[,"fit"])), big.mark=","),
  CI_width_DKK = format(round(ci_widths), big.mark=","),
  PI_width_DKK = format(round(pi_widths), big.mark=","),
  PI_CI_ratio  = round(pi_widths / ci_widths, 1)
)
knitr::kable(width_tbl,
             col.names = c("sqm","Point est (DKK)",
                           "CI width (DKK)","PI width (DKK)","PI/CI ratio"),
             caption   = "Interval widths on DKK scale")
```

---

# Plot: Predicted Price vs sqm

```{r predict-plot}
#| results: hold


sqm_seq <- seq(60, 250, by = 2)
new_seq <- data.frame(
  log_sqm            = log(sqm_seq),
  house_age          = 55,
  nom_interest_rate. = 3.5,
  dk_ann_infl_rate.  = 3.3
)

ci_seq <- predict(M_final, newdata = new_seq, interval = "confidence")
pi_seq <- predict(M_final, newdata = new_seq, interval = "prediction")

ylim_range <- range(exp(pi_seq[, c("lwr","upr")]))

plot(sqm_seq, exp(ci_seq[,"fit"]),
     type = "l", lwd = 2.5, col = "steelblue",
     ylim = ylim_range,
     main = "Predicted Price vs sqm — Copenhagen Villa\n(house_age=55, nom_rate=3.5%, infl=3.3%)",
     xlab = "sqm",
     ylab = "Price (DKK)")

# CI band
polygon(c(sqm_seq, rev(sqm_seq)),
        c(exp(ci_seq[,"lwr"]), rev(exp(ci_seq[,"upr"]))),
        col = rgb(0.27, 0.51, 0.71, 0.2), border = NA)

# PI band
polygon(c(sqm_seq, rev(sqm_seq)),
        c(exp(pi_seq[,"lwr"]), rev(exp(pi_seq[,"upr"]))),
        col = rgb(0.70, 0.13, 0.13, 0.1), border = NA)

lines(sqm_seq, exp(ci_seq[,"lwr"]), lty = 2, col = "steelblue", lwd = 1.5)
lines(sqm_seq, exp(ci_seq[,"upr"]), lty = 2, col = "steelblue", lwd = 1.5)
lines(sqm_seq, exp(pi_seq[,"lwr"]), lty = 2, col = "firebrick", lwd = 1.5)
lines(sqm_seq, exp(pi_seq[,"upr"]), lty = 2, col = "firebrick", lwd = 1.5)

# Mark the three new houses
points(exp(new_houses$log_sqm), exp(ci_log[,"fit"]),
       pch = 16, cex = 1.4, col = "black")

legend("topleft",
       legend = c("Point estimate",
                  "95% CI for mean (shaded)",
                  "95% PI for single house (shaded)"),
       col    = c("steelblue","steelblue","firebrick"),
       lty    = c(1, 2, 2), lwd  = c(2.5, 1.5, 1.5),
       bty    = "n", cex = 0.85)
```

---

# Sensitivity to Interest Rate

How does the predicted price change as `nom_interest_rate.` varies?

```{r sensitivity}
#| results: hold


rates   <- seq(1.75, 3.6, by = 0.25)
base_df <- data.frame(log_sqm            = log(130),
                       house_age          = 55,
                       nom_interest_rate. = rates,
                       dk_ann_infl_rate.  = 3.3)

preds <- predict(M_final, newdata = base_df, interval = "confidence")

plot(rates, exp(preds[,"fit"]) / 1e6,
     type = "b", pch = 16, col = "steelblue", lwd = 2,
     main = "Predicted Price vs Nominal Interest Rate\n(130 sqm, age=55, infl=3.3%)",
     xlab = "Nominal interest rate (%)",
     ylab = "Predicted price (million DKK)")
lines(rates, exp(preds[,"lwr"]) / 1e6, lty = 2, col = "steelblue")
lines(rates, exp(preds[,"upr"]) / 1e6, lty = 2, col = "steelblue")

cat("Price at 1.75% rate:", format(round(exp(preds[1,"fit"])), big.mark=","), "DKK\n")
cat("Price at 3.60% rate:", format(round(exp(preds[nrow(preds),"fit"])), big.mark=","), "DKK\n")
cat("Difference:         ", format(round(exp(preds[1,"fit"]) -
                                         exp(preds[nrow(preds),"fit"])),
                                   big.mark=","), "DKK\n")
```

---

# Interpretation

**CI vs PI — the key distinction:**

A CI for the mean answers: *"What is the typical price of all 130 sqm villas with these characteristics?"* The CI narrows as $n \to \infty$ because we learn the mean precisely.

A PI for a new house answers: *"What will this specific house sell for?"*
The PI never collapses to a point because individual transactions always
carry irreducible noise $\sigma^2$, no matter how large the dataset.

**Back-transformation note:** the intervals are symmetric on the log scale but asymmetric on the DKK scale (wider above the point estimate than below), correctly reflecting the log-normal structure of prices.
