---
title: "Step 5: Comparison of Regression Lines (ANCOVA)"
subtitle: "Danish Housing Prices · Mathematical Statistics"
author: "Carlos Castillo"
date: today
format:
  pdf:
    pdf-engine: pdflatex  # El motor más rápido para PDF
    papersize: a4         # Formato estándar en Dinamarca
    number-sections: true
    keep-tex: false       # No pierde tiempo guardando archivos basura
execute:
  cache: true             # Fundamental para velocidad en R
  freeze: auto            # Evita re-renders innecesarios
  echo: true              # Muestra tu código para el profesor
  warning: false
  results: hold           # Agrupa resultados para un renderizado más limpio
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.height = 4.5, fig.width = 8)
```

```{r load}
#| results: hold


library(tidyverse)
dk_clean <- readRDS("../../data/dk_clean.rds")

types  <- levels(dk_clean$house_type)
colors <- c("steelblue","firebrick","forestgreen","darkorange")

cat("House types:", types, "\n")
cat("n =", nrow(dk_clean), "\n")
```

---

# Question

Does $\log(\text{sqm})$ have the **same slope** for all house types, or does
each type scale differently with size?

---

# Three-Model Hierarchy

Let $t_i = \log(\text{sqm}_i)$ and let $j(i)$ denote the house type of
observation $i$:

**M1** (separate intercepts and slopes):
$$Y_i \sim \mathcal{N}(\alpha_{j(i)} + \beta_{j(i)}\, t_i,\; \sigma^2)$$

**M2** (separate intercepts, common slope — ANCOVA):
$$Y_i \sim \mathcal{N}(\alpha_{j(i)} + \beta\, t_i,\; \sigma^2)$$

**M3** (single regression line):
$$Y_i \sim \mathcal{N}(\alpha + \beta\, t_i,\; \sigma^2)$$

---

# Variance Homogeneity: Bartlett's Test on Residuals

**Why residuals, not the raw response?**
Under a regression model, variation in $Y$ is partly explained by $t_i$.
Testing on raw $Y$ would confound structural variance with regression signal.
The slides (§8) show the correct syntax: pass the **fitted `lm` objects**
directly to `bartlett.test`. R then uses the residual variance estimates
$s_{(i)}^2 \sim \sigma_i^2 \chi^2(n_i-2)/(n_i-2)$.

$$H_{0\sigma^2}: \sigma_{\text{Apt}}^2 = \sigma_{\text{Sum}}^2
= \sigma_{\text{Town}}^2 = \sigma_{\text{Villa}}^2$$

```{r bartlett-residuals}
#| results: hold


# Fit separate regression within each house type (= the M1 model)
fits_list <- split(dk_clean, dk_clean$house_type)
fits_list <- lapply(fits_list, function(d) lm(log_price ~ log_sqm, data = d))

cat("Residual SDs from group-specific regressions:\n")
for (tp in types) {
  cat(" ", tp, ":", round(summary(fits_list[[tp]])$sigma, 4), "\n")
}

cat("\nMax / Min residual SD ratio:",
    round(max(sapply(fits_list, function(f) summary(f)$sigma)) /
          min(sapply(fits_list, function(f) summary(f)$sigma)), 2), "\n\n")

# Correct call: pass lm objects, not raw response
cat("--- Bartlett's test (on regression residuals) ---\n")
bartlett.test(fits_list)
```

---

# Exploratory Plot: Separate Regression Lines

```{r scatter-lines}
#| results: hold


plot(dk_clean$log_sqm, dk_clean$log_price,
     pch = 16, cex = 0.08, col = rgb(0, 0, 0, 0.07),
     main = "log(Price) vs log(sqm) by House Type",
     xlab = "log(sqm)", ylab = "log(Price)")

for (j in seq_along(types)) {
  abline(fits_list[[types[j]]], col = colors[j], lwd = 2)
}
legend("topleft", legend = types, col = colors,
       lwd = 2, bty = "n", cex = 0.8)
```

---

# Model Tests

## Test 1: Common slope? (M2 vs M1)

$$H_0: \beta_1 = \beta_2 = \beta_3 = \beta_4 = \beta$$

```{r m1-m2}
#| results: hold


M1_reg <- lm(log_price ~ house_type * log_sqm, data = dk_clean)
M2_reg <- lm(log_price ~ house_type + log_sqm, data = dk_clean)

cat("--- F-test: M2 vs M1 (common slope) ---\n")
anova(M2_reg, M1_reg)

cat("\ns(M1) =", round(summary(M1_reg)$sigma, 5), "\n")
cat("s(M2) =", round(summary(M2_reg)$sigma, 5), "\n")
```

## Test 2: No house type effect? (M3 vs M2)

$$H_0: \alpha_1 = \alpha_2 = \alpha_3 = \alpha_4$$

```{r m2-m3}
#| results: hold


M3_reg <- lm(log_price ~ log_sqm, data = dk_clean)

cat("--- F-test: M3 vs M2 (no type effect) ---\n")
anova(M3_reg, M2_reg)

cat("\ns(M3) =", round(summary(M3_reg)$sigma, 5), "\n")
```

---

# Slope Estimates per House Type

Using the no-intercept parameterisation to obtain individual slope estimates:

```{r slopes}
#| results: hold


M1_slopes <- lm(log_price ~ 0 + house_type + house_type:log_sqm,
                data = dk_clean)

cf <- summary(M1_slopes)$coefficients
slope_rows <- grep(":log_sqm", rownames(cf))

cat("Slope estimates (beta_j) with 95% CIs:\n\n")
slopes_ci <- confint(M1_slopes)[slope_rows, ]
slopes_est <- cf[slope_rows, c("Estimate","Std. Error","t value","Pr(>|t|)")]

print(round(slopes_est, 4))
cat("\n95% CIs for slopes:\n")
print(round(slopes_ci, 4))
```

---

# ANCOVA Model (M2): Summary

If the common-slope test is not rejected, M2 is the working model.
It gives a single slope $\hat\beta$ (shared price-size elasticity) and
separate intercepts per house type (price-level differences).

```{r m2-summary}
#| results: hold


summary(M2_reg)
cat("\nInterpretation of slope: a 1% increase in sqm is associated with\n")
cat("a", round(summary(M2_reg)$coefficients["log_sqm","Estimate"] * 100, 2),
    "basis-point increase in log-price (approx",
    round(summary(M2_reg)$coefficients["log_sqm","Estimate"], 3),
    "% increase in price).\n")
```

---

# Residual Diagnostics: M2

```{r diag-m2}
#| results: hold


par(mfrow = c(1, 2))

res_m2 <- rstandard(M2_reg)

plot(fitted(M2_reg), res_m2,
     pch = 16, cex = 0.1, col = rgb(0,0,0.6,0.15),
     main = "Residuals vs Fitted — M2",
     xlab = "Fitted", ylab = "Std. residuals")
abline(h =  0, lty = 1)
abline(h =  2, lty = 2, col = "firebrick")
abline(h = -2, lty = 2, col = "firebrick")

qqnorm(res_m2, pch = 16, cex = 0.1, col = "steelblue",
       main = "QQ — M2 residuals")
qqline(res_m2, col = "firebrick", lwd = 2)

par(mfrow = c(1, 1))
```
