---
title: "Step 3: k-Sample ANOVA (House Type Effect)"
subtitle: "Danish Housing Prices · Mathematical Statistics"
author: "Carlos Castillo"
date: today
format:
  pdf:
    pdf-engine: pdflatex  # El motor más rápido para PDF
    papersize: a4         # Formato estándar en Dinamarca
    number-sections: true
    keep-tex: false       # No pierde tiempo guardando archivos basura
execute:
  cache: true             # Fundamental para velocidad en R
  freeze: auto            # Evita re-renders innecesarios
  echo: true              # Muestra tu código para el profesor
  warning: false
  results: hold           # Agrupa resultados para un renderizado más limpio
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.height = 4, fig.width = 7)
```

```{r load}
#| results: hold


library(tidyverse)
dk_clean <- readRDS("../../data/dk_clean.rds")

cat("Group sizes:\n")
print(table(dk_clean$house_type))
```

---

# Statistical Model

Let $j$ index house type ($j = 1, \ldots, 4$) and $i$ index observations
within each group:

$$Y_{ij} \sim \mathcal{N}(\mu_j, \sigma^2),
\quad j = 1,\ldots,4,\quad i = 1,\ldots,n_j
\quad \text{(independent)}$$

The equal-variance assumption $\sigma^2$ common across groups must be
checked before pooling.

---

# Variance Homogeneity — Bartlett's Test

We must perform **Bartlett's test** for $k \geq 3$ groups, applied to the response grouped by factor levels:

$$H_{0\sigma^2}: \sigma_1^2 = \sigma_2^2 = \sigma_3^2 = \sigma_4^2$$

$$\text{Ba} = \frac{-2 \ln Q}{C} \sim \chi^2(k-1) \quad \text{under } H_{0\sigma^2}$$

where $C = 1 + \frac{1}{3(k-1)}\!\left(\sum_{i=1}^k \frac{1}{f_{(i)}} - \frac{1}{f_1}\right)$
is Bartlett's correction factor.

```{r bartlett}
#| results: hold


cat("Per-group standard deviations:\n")
print(round(tapply(dk_clean$log_price, dk_clean$house_type, sd), 4))

cat("\n--- Bartlett's test ---\n")
bartlett.test(log_price ~ house_type, data = dk_clean)

cat("\nNote: with n ~ 83,000 Bartlett has near-infinite power.\n")
cat("Inspect the SDs directly: if max/min ratio < 2, equal variance\n")
cat("is a reasonable working assumption.\n")
```

---

# Exploratory Plots

```{r plots}
#| results: hold


par(mfrow = c(1, 2))

boxplot(log_price ~ house_type, data = dk_clean,
        col  = c("lightsteelblue","lightcoral","lightgreen","lightyellow"),
        main = "log(Price) by House Type",
        xlab = "", ylab = "log(Price)", las = 2)

# Per-group histograms overlaid
types  <- levels(dk_clean$house_type)
colors <- c("steelblue","firebrick","forestgreen","darkorange")
xlim   <- range(dk_clean$log_price)

plot(NULL, xlim = xlim, ylim = c(0, 2.5),
     main = "Density by House Type",
     xlab = "log(Price)", ylab = "Density")
for (j in seq_along(types)) {
  lines(density(dk_clean$log_price[dk_clean$house_type == types[j]]),
        col = colors[j], lwd = 2)
}
legend("topright", legend = types, col = colors, lwd = 2, bty = "n", cex = 0.7)

par(mfrow = c(1, 1))
```

---

# k-Sample ANOVA — F-Test

## Hypothesis

$$H_0: \mu_1 = \mu_2 = \mu_3 = \mu_4
\quad \text{vs} \quad
H_1: \text{at least one } \mu_j \text{ differs}$$

## F-statistic

$$F = \frac{(\text{RSS}_{M_\infty} - \text{RSS}_{M_1})/(k-1)}
           {\text{RSS}_{M_1}/(n-k)}
\sim F(k-1,\, n-k) \quad \text{under } H_0$$

```{r anova-models}
#| results: hold


M_full <- lm(log_price ~ house_type, data = dk_clean)  # group-means model
M_null <- lm(log_price ~ 1,          data = dk_clean)  # common-mean model

cat("--- ANOVA table (full model first per course convention) ---\n")
anova(M_null, M_full)
```

```{r anova-summary}
summary(M_full)
```

---

# Group Estimates

```{r estimates}
#| results: hold


cat("Group means on log scale:\n")
grp_means <- tapply(dk_clean$log_price, dk_clean$house_type, mean)
print(round(grp_means, 4))

cat("\nGroup medians on DKK scale (exp of mean log-price):\n")
print(format(round(exp(grp_means)), big.mark = ","))

cat("\nPooled variance estimate s^2:\n")
cat(round(summary(M_full)$sigma^2, 6), "\n")
cat("s =", round(summary(M_full)$sigma, 4), "\n")
```

---

# Confidence Intervals for Group Means

```{r cis}
#| results: hold


ci_mat <- confint(M_full)
cat("95% CIs for model parameters (reference = Apartment):\n")
print(round(ci_mat, 4))
```

---

# Interpretation

The F-test compares the **group-means model** $M_1$ (4 free means) to
the **common-mean model** $M_\infty$ (1 mean). Rejection means at least
one house type has a different mean log-price — but not which ones.

The `summary()` output shows pairwise contrasts against the reference
level (Apartment by default in R). Each t-test there tests
$H_0: \mu_j - \mu_\text{Apartment} = 0$.
